Using an SMT engine to generate Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-20 2017-Xudong-SMT/ActaInformatica.tex
lot of text in section "smart"

In Fig.~\ref{schema:EnableEncodings}, we show examples of these
constructs, with two pLTSs, one hole and one pNet node encoding our running example.

\subsection{Example}
\label{section:example}
\TODO{Taken from \cite[Forte]}

@@

        In this example we use \emph{synchronised
          actions}, to express actions that cannot be further synchronised, like the
        $\tau$ silent action of CCS. We denote them as any action expression with the text
        underlined, as e.g. $\underline{\delta(x_2)}$. Such
        synchronised actions do not play any special role for defining
        the semantics of pNets, but as one can expect, will be crucial
        for defining weak equivalences.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a pLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State \TODO{Xudong:Prune the unsatisfiable open transitions in $L$ using the SMT
  solver;
   and add a call to ``compute the reachable sub-automaton'';}
\State \Return $\mathit{OA}=(\mathit{InitState}(P),L)$;

@@

\ERIC{Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instanciations. This is not logicaly
incorrect, but is naturaly not an optimal representation. So we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions, and checked for satisfiability.\\
Then we compute the set of reachable transitions and states to build
the final open automaton.}
%By construction, due to the
%satisfiability test in line 6, it only produces the set of states reachable
%through a sequence of satisfiable open transitions.

@@

In a pLTS, there may be several incoming transitions 
that assign potentially different (symbolic) values to a state variable.
To handle such cases, the algorithm manages, for each pLTS state, a
list of expressions collected from the assignments of each state
variable. %% $v$.
%% The list we used contains triples <$v$, $S$, $AssignRH$> 
%% where $v$ is the variable in pLTS, $S$ is its owner state and
%% $AssignRH$ is a list of expressions over other variables in the right
%% hand side of assignments of $v$.
% As a pragmatic extension to the formal definition, we also manage
% ``global variables'', defined in all states of the pLTS.
For a global state in the open automaton, the set of state
variables (which may be used in an outgoing transition) %% will simply be
is the disjoint union
of sets of state variables of the individual pLTS states constituting this
global state.
\TODO{Xudong: speciy more precisely what is done, and at which point
  in the MakeTransitions algo}
%There are some assignment on the variables in each of states of the pLTS (state variables), or just as global states of it (global variables). There is usually not only one assignment for a state variable, miss of them may drop the unsatisfiable results. In order to avoid such mistakes, 
\TODO{Eric: these sets of assignment collect potential assignments from all OTs
  in the generated automaton. If we prune the generated state
  space, we will get less assignements, and potentialy also get better
  minimization... May be move this to ``smart'' section}

@@

\subsection{Computing reachable state space on the fly}
\TODO{add a high level description of the new strategy, discuss
  decidability/approximation of the SAT checking, express some
  ``minimality'' property}

The basic algorithm from the previous section potentialy explores some
areas of the global state-space that will not be reachable after
checking the OTs' predicates for satisfiability. Incorporating the
satisfiability check inside the residual algorithm will definitely
save some of these useless computation. More precisely, it will still
create, and submit to Z3, the OTs at the ``border'' of the state-space, but
not the transitions issued from non-reachable global states.

\ERIC{But there is more to this. As mentionned above, we collect all
possible (symbolic) assignments of state variables, in the
computed open automaton. By avoiding building the OTs coming from
unreachable states, we also avoid adding the corresponding assignments
to the variables of their target states. The set of possible
assignments is an over approximation of possible values of the
variables of such state, so narrowing it means we get a better
approximation, and potentialy more informations, that can be used to
the SAT check of subsequent transitions. But we \emph{cannot} use this
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. So the smart algorithm performs in to steps: during the
first step, the SAT check is done without using any assignment set
information. Then after completion of this step, we perform another
SAT check of all transitions from step1, but this time using the
assignment information. This may reduce the reachable state space. If
we want to be sure the obtained automaton is minimal, we should
iterate this step until a fix-point is reached. But in practical cases, very
little will be gained, and leaving some UNSAT transitions in the
symbolic automaton is still logicaly correct.}

Algorithm~\ref{alg3} implement this new stategy: \TODO{Explain
  significant detials}

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
%%        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
%%\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\State \Return $\mathit{OA}=(\mathit{InitState}(P),L)$;

@@

\ERIC{But in presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still the smart strategy will explore a reduced state
  space, but not the minimal one (discussion later)}

@@

\TODO{Eric : provide counter-example for smart with unsing incomplete
  assignment sets}

>>>>

In Figures~\ref{schema:enable-pnets},\ref{schema:enable-composed}, we show
examples of these constructs, with two pLTSs, one hole and one pNet
node encoding our running example.

\subsection{Example}
\label{section:example}

@@

        In this example we use \emph{synchronised
          actions}, to express actions that cannot be further synchronised, likesimilar to the
        $\tau$ silent action of CCS. We denote them as any action expression with the text
        underlined, as e.g. $\underline{\delta(x_2)}$. Such
        synchronised actions do not play any special role for defining
        the semantics of pNets, but as one can expect, will be crucial
        for defining weak equivalences.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a pLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State \QIN{$\mathit{L1}$ = \texttt{filterUnsatTransitions}($\mathit{L,Assignements}$)}
\State \QIN{$\mathit{OA}$ = \texttt{makeReachableSubAutomaton}($\mathit{InitState}(P)$,$\mathit{L1}$)}
\State \Return $\mathit{OA}$;

@@

\ERIC{Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instanciations. This is not logicaly
incorrect, but is naturaly not an optimal representation. So we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions, and checked for satisfiability (function \texttt{filterUnsatTransitions}).\\
Then we compute the set of reachable transitions and states to build
the final open automaton.}

@@

In a pLTS, there may be several incoming transitions 
that assign potentially different (symbolic) values to a state variable.
To handle such cases, the algorithm manages a
list of assignment expressions for each pLTS state. 
\QIN{The only assignments happen on state variables, at pLTS level: every time the MakeTransitions
algorithm is applied on a pLTS, it collects the expressions from the $Post$ in each transition.
% As there is no assignments collecting work on a pNet node, it is not
% shown in Algorithm~\ref{alg2}.
}

For a global state in the open automaton, the set of state
variables (which may be used in an outgoing transition) 
is the disjoint union
of sets of state variables of the individual pLTS states constituting this
global state. \ERIC{The final assignment sets are assembled when building the
open automaton from the reachable transitions.}

@@

\subsection{Computing reachable state space on the fly}

\ERIC{The basic algorithm from the previous section potentialy explores some
areas of the global state-space that will not be reachable after
checking the OTs' predicates for satisfiability. Incorporating the
satisfiability check inside the residual algorithm will definitely
save some of these useless computation. More precisely, it will still
create, and submit to Z3, the OTs at the ``border'' of the state-space, but
not the transitions issued from non-reachable global states. We use
this idea to create our ``smart'' algorithm.}

But there is more to this. Due to the symbolic nature of the
  transitions, reachability is not as easy to define as in standard
  models. As mentionned above, we collect all
possible (symbolic) assignments of state variables, in the
computed open automaton. By avoiding building the OTs coming from
some unreachable states, we also avoid adding the corresponding assignments
to the variables of their target states. The set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentialy more information, that can be used to
the SAT checking of subsequent transitions. But we \emph{cannot} use this
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. So the smart algorithm performs in two steps: during the
first step, the SAT checking is done without using any assignment set
information. Then after completion of this step, we perform another
SAT checking of all transitions from step1, but this time using the
assignment information. This may reduce the reachable state space. If
we want to be sure the obtained automaton is minimal, we should
iterate this step until a fix-point is reached. But in practical cases, very
little will be gained, so the default implementation does only one
step. Leaving some UNSAT transitions in the
symbolic automaton is still logicaly correct.

\TODO{Yuxin: I can add an example showing in detail why we cannot use
  the assignment information on the fly. Do you think this is useful,
  or id the explanation above sufficient ?}

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
%%        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
%%\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\IfThenElse {(NonEmptyAssignmentSet($\mathit{L}$))}
        {\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
          \State \Return $\mathit{OA}$=makeAutomaton($\mathit{InitState}(P),L)$;}
        {\State \Return $\mathit{OA}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$};
\end{algorithmic}  
\end{algorithm}

Algorithm~\ref{alg3} implement this new stategy: in line 5, it runs
the satisfiability check on each OT generated from this global
state. This check is done without using the assignment information, as
discussed above. Only the OTs found SAT from this check will be added
to $L$, and the corresponding target global state added to $U$ if
necessary. This ensures that all explored global states are
effectively reachable by a chain of SAT OTs.

Then at the end of the residual loop, we do another traversal of the
automaton, this time using the assignment information collected that
far, and compute the final reachable sub-automaton. This step,
naturaly, is only useful if there are some assignments.

@@

\ERIC{But in presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still, the smart strategy will explore a reduced state
  space, but not the minimal one (discussion later)}

@@

\TODO{Eric : Eventually provide counter-example for smart with unsing incomplete
  assignment sets}
